---
title: "Upwelling GLM - Seasonal Means"
author: "Gammon Koval"
date: "2022-12-13"
output:
  html_document: default
editor_options:
  chunk_output_type: console
---

# Script 6 - Assess GLM on the Seasonal SST Footprint Data 

```{r libraries, results = 'hide', warning = FALSE, message = FALSE, error = FALSE}
#load libraries
library(tidyverse)
library(lubridate)
library(ggpubr)
library(rstatix)
library(broom)
library(car)
library(nlme)   
library(gridExtra)
library(emmeans)
library(multcomp)
library(MASS)
library(interactions)
library(lmtest)
library(knitr)
library(kableExtra)

rm(list = ls())
```

## Data Preparation and Exploration 

```{r data-prep-regular, out.width = '95%'}
#read in data
setwd('./Documents/GitHub/Garcia-Reyes_etal_2023_EBUSFootprint')
upwelling <- read_csv('./data/upwellingfootprint_seasonal_nearshore_Modified-Subregions.csv')

#separate the Random_name variable on the - 
upwelling <- upwelling %>% 
  separate(Random_name, c('EBUS', 'Subregion'), '-')

#replace the '-' in Iberian-Canary with '/'
upwelling$EBUS <- gsub("-", "/", upwelling$EBUS) 
  
#add back in random factor
upwelling$Random <- paste(upwelling$EBUS, upwelling$Subregion, sep = '-')

#create a column of named seasons
upwelling$Season <- NA
for (i in 1:length(upwelling$Season_month)) {
  if (upwelling$EBUS[i] %in% c('California', 'Iberian/Canary')) {
    if (upwelling$Season_month[i] == 6){
      upwelling[i,]$Season = 'Summer'
    } else if (upwelling$Season_month[i] == 9){
      upwelling[i,]$Season = 'Fall'
    } else if (upwelling$Season_month[i] == 12){
      upwelling[i,]$Season = 'Winter'
    } else {
      upwelling[i,]$Season = 'Spring'
    }
  } else {
    if (upwelling$Season_month[i] == 6){
      upwelling[i,]$Season = 'Winter'
    } else if (upwelling$Season_month[i] == 9){
      upwelling[i,]$Season = 'Spring'
    } else if (upwelling$Season_month[i] == 12){
      upwelling[i,]$Season = 'Summer'
    } else {
      upwelling[i,]$Season = 'Fall'
    }
  }
}

#relabel EBUS factors
upwelling$EBUS <- factor(upwelling$EBUS, levels = c('California', 'Humboldt', 'Iberian/Canary', 'Benguela'))
upwelling$Subregion <- factor(upwelling$Subregion, levels = c('Equatorward', 'Central', 'Poleward'))
upwelling$Season <- factor(upwelling$Season, levels = c('Winter', 'Spring', 'Summer', 'Fall'))

#plot year vs percent grouped by EBUS and subregion
ggplot(upwelling, aes(x = Year, y = Percent, group = Season, color = Season)) + 
    geom_point() + 
    facet_wrap(EBUS~Subregion, ncol = 3, scales = 'free_y') +
    geom_smooth(method = 'lm', se = FALSE) + 
    theme_bw()

#plot percent data grouped by EBUS and subregion
ggplot(upwelling, aes(x = Percent, group = Season, color = Season)) +
  geom_density() + 
  facet_wrap(EBUS ~ Subregion, ncol = 3, scales = 'free_y')

```

## Run the Model 

The full model includes seven variables: 

* 4 single interactions: Year, EBUS, Subregion, and Season
* 6 two-way interactions: Year & EBUS, Year & Subregion, EBUS & Subregion, Year & Season, EBUS & Season, and Subregion & Season 
* 4 three-way interactions: Year & EBUS & Subregion, Year & EBUS & Season, Year & Subregion & Season, and EBUS & Subregion & Season
* 1 four-way interaction: Year & EBUS & Subregion & Season

```{r final-model-regular}
#run the final model
MFinal <- lm(Percent ~ Year * EBUS * Subregion * Season, data = upwelling)

anova(MFinal)

#view model output
summary(MFinal)
```

The full model is significant (p < 0.05) with an adjusted r-squared of 0.06. Year, Year:EBUS, and Year:Subregion were all significant with Year:Season being slightly significant (p < 0.01). 

## Model Diagnostics

```{r diagnostics-regular, out.width = '95%'}
#look at residuals
par(mfrow = c(2,2))
plot(MFinal)

#check residuals for normality (p < 0.05 means the residuals aren't normal)
shapiro_test(MFinal$residuals)

#check for auto-correlation
par(mfrow = c(1,1))
acf(MFinal$residuals)

#Breusch-Godfrey Test (p < 0.05 means there is autocorrelation)
bgtest(upwelling$Percent~upwelling$Year, order=1)

ggplot(data = upwelling, aes(x = MFinal$residuals)) +
    geom_histogram(aes(y = ..density..), fill = 'steelblue', color = 'black') +
    labs(title = 'Histogram of Residuals', x = 'Residuals', y = 'Density') + 
    stat_function(fun = dnorm, args = list(mean = mean(MFinal$residuals), sd = sd(MFinal$residuals)), color = 'navy', size = 1, linetype = 'longdash')
```

The model seem to fit the data fine. It doesn't pass the Shapiro normality test, but looking at the histogram it is very normal, but there is a strong amount of autocorrelation. Though I'm not sure how this if factored in now that season is included in the model. 

## Post Hoc Tests 

### Four-Way Interaction

```{r four-way-interaction, out.width = '95%'}
#extracts trends and p-values from the models
upwelling.emt <- as.data.frame(emtrends(MFinal, pairwise ~ EBUS * Subregion * Season, var = "Year")$emtrends)

upwelling.test <- as.data.frame(test(emtrends(MFinal, pairwise ~ EBUS * Subregion * Season, var = "Year"))$emtrends)

upwelling.posthoc <- dplyr::full_join(upwelling.emt, upwelling.test, by = c("EBUS", "Subregion", "Season", "Year.trend", 'SE', 'df'))

#drop NAs
upwelling.posthoc <- upwelling.posthoc[complete.cases(upwelling.posthoc), ]
write.csv(upwelling.posthoc, './data/seasonal_trends_percent_Modified-Subregions_new.csv', row.names = FALSE)

#view data
upwelling.posthoc %>% 
  kbl() %>%
  kable_styling()

#Order EBUS and Subregion factors
upwelling.posthoc$EBUS <- factor(upwelling.posthoc$EBUS, levels = c('California', 'Humboldt', 'Iberian/Canary', 'Benguela'))
upwelling.posthoc$Subregion <- factor(upwelling.posthoc$Subregion, levels = c('Equatorward', 'Central', 'Poleward'))

#plot trends with CIs
ggplot(upwelling.posthoc, aes(x = Season, y = Year.trend)) + 
  geom_errorbar(aes(ymin=lower.CL, ymax=upper.CL), width=.2,
               position=position_dodge(.9)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = 'dashed', color = 'grey60') +
  theme_bw() +
  facet_wrap(EBUS ~ Subregion, ncol = 3) + 
  ylab('Trend ± 95% CI') +
  ggtitle('Trends for Season Seasonal Means')
```

