---
title: "Upwelling GLM - Annual"
author: "Gammon Koval"
date: "2022-11-11"
output:
  word_document: default
  html_document: default
editor_options:
  chunk_output_type: console
---

# Script 5 - Assess GLS on Annual Data with Nearshore Cutoff

Using the GLS function, we can build in an ARMA autocorrelation structure to the model.

```{r libraries, results = 'hide', warning = FALSE, message = FALSE, error = FALSE}
#load libraries
library(tidyverse)
library(ggpubr)
library(rstatix)
library(broom)
library(car)
library(nlme)   
library(gridExtra)
library(emmeans)
library(multcomp)
library(MASS)
library(interactions)
library(lmtest)
library(orcutt)

rm(list = ls())
```

## Annual Data

### Data Preparation and Exploration 

```{r data prep}
#read in data
setwd('./Documents/GitHub/Garcia-Reyes_etal_2023_EBUSFootprint')
upwelling <- read_csv('./data/upwellingfootprint_annualsummary_nearshore_Modified-Subregions.csv')

#add columns for transforming the data and separate Region back into EBUS and subregion
upwelling <- upwelling %>% 
  separate(Random_name, into = c('EBUS', 'Subregion'), sep = '-')

#add back in random factor
upwelling$Random <- paste(upwelling$EBUS, upwelling$Subregion, sep = '-')

#relabel EBUS factors
upwelling$EBUS <- factor(upwelling$EBUS, levels = c('California', 'Humboldt', 'Iberian/Canary', 'Benguela'))
upwelling$Subregion <- factor(upwelling$Subregion, levels = c('Equatorward', 'Central', 'Poleward'))

#plot year vs percent grouped by EBUS and subregion
ggplot(upwelling, aes(x = Year, y = Percent)) + 
    geom_point() + 
    facet_wrap(EBUS~Subregion, ncol = 3, scales = 'free_y') +
    geom_smooth(method = 'lm', se = FALSE) + 
    theme_bw()

#plot percent data grouped by EBUS and subregion
ggplot(upwelling, aes(x = Percent)) +
  geom_density() + 
  facet_wrap(EBUS ~ Subregion, ncol = 3, scales = 'free_y')

```

### Base Model 

```{r base-model}
#run a linear model
Mlm <- lm(Percent ~ Year * EBUS * Subregion, 
             data = upwelling)

#run the gls model with variance partitioned
MBase <- gls(Percent ~ Year * EBUS * Subregion, 
             weights = varIdent(form = ~ 1 | EBUS),
             method = 'REML',
             data = upwelling)

#view model output
anova(MBase)
summary(MBase) #AIC 1586.597

#look at residuals
par(mfrow = c(2,2))
plot(MBase)

#another way to look at the residuals
colvec <- c("#ff1111","#007eff") ## second colour matches lattice default
grid.arrange(plot(MBase,type=c("p","smooth")),
             plot(MBase,sqrt(abs(resid(.)))~fitted(.),
                  type=c("p","smooth"),ylab=expression(sqrt(abs(resid)))),
                                       ## "sqrt(abs(resid(x)))"),
             plot(MBase,resid(.,type="pearson")~Year,
                  type=c("p","smooth")),
             qqnorm(MBase,abline=c(0,1)))


#check residuals for normality (p < 0.05 means the residuals aren't normal)
shapiro_test(MBase$residuals)

#check for auto-correlation
par(mfrow = c(1,1))
acf(MBase$residuals)

#Breusch-Godfrey Test (p < 0.05 means there is autocorrelation)
bgtest(upwelling$Percent~upwelling$Year, order=1)
```

### Testing Correlation Structures

```{r testing-correlation}
#loop through a list of p's 
for (q in 0:3) {
  #loop through a list of q's
  for (p in 0:3) {
    #skip 0,0
    if ((p == 0 & q == 0) | (p == 1 & q == 2)| (p == 1 & q == 3)) {
      next
    } else {
      #test the model with the various ARMA parameters
      MARMA <- gls(Percent ~ Year * EBUS * Subregion, 
                   weights = varIdent(form = ~ 1 | EBUS),
                   correlation = corARMA(form = ~ Year | EBUS/Subregion, 
                                         p = p, q = q),
                   method = 'REML',
                   data = upwelling)
      
      #print the autocorrelation
      print(paste(p, q, round(BIC(MARMA), 2)))
    }
  }
}

```

Based on the BICs, ARMA(1, 0) is the best model. 

### Best Model

#### Run Model

```{r best-model}
#run the final model
MFinal <- gls(Percent ~ Year * EBUS * Subregion,
              weights = varIdent(form = ~ 1 | EBUS),
              correlation = corARMA(form = ~ Year | EBUS/Subregion, p = 1, q = 0),
              method = 'REML',
              data = upwelling)

#view model output
anova(MFinal)
summary(MFinal) #AIC 1523.395
```

#### Look at Residuals

```{r best-residuals}
#look at residuals
par(mfrow = c(2,2))
plot(MFinal)

#another way to look at the residuals
colvec <- c("#ff1111","#007eff") ## second colour matches lattice default
grid.arrange(plot(MFinal,type=c("p","smooth")),
             plot(MFinal,sqrt(abs(resid(.)))~fitted(.),
                  type=c("p","smooth"),ylab=expression(sqrt(abs(resid)))),
                                       ## "sqrt(abs(resid(x)))"),
             plot(MFinal,resid(.,type="pearson")~Year,
                  type=c("p","smooth")),
             qqnorm(MFinal,abline=c(0,1)))


#check residuals for normality (p < 0.05 means the residuals aren't normal)
shapiro_test(MFinal$residuals)

#check for auto-correlation
par(mfrow = c(1,1))
acf(residuals(MBase, type = 'normalized'))
```


#### Post Hoc Tests 

The following code outputs the slopes of the trends for EBUS * Subregion. It is also possible to do additional post-hoc tests, but since we don't care about that, I didn't include those trends.

```{r best-post-hoc}
# emmeans_pwc = emmeans(MFinal, 'EBUS', by = 'Subregion', adjust = 'mvt')
# summary(emmeans_pwc)

#extracts trends from the linear model
upwelling.emt <- as.data.frame(emtrends(Mlm, pairwise ~ EBUS * Subregion, var = "Year", mode = "df.error")$emtrends)

upwelling.test <- as.data.frame(test(emtrends(Mlm, pairwise ~ EBUS * Subregion, var = "Year", mode =  "df.error"))$emtrends)

upwelling.posthoc.lm <- dplyr::full_join(upwelling.emt, upwelling.test, by = c("EBUS", "Subregion", "Year.trend", 'SE', 'df'))

upwelling.posthoc.lm$Model = 'lm'

#extracts trends from the base model
upwelling.emt <- as.data.frame(emtrends(MBase, pairwise ~ EBUS * Subregion, var = "Year", mode = "df.error")$emtrends)

upwelling.test <- as.data.frame(test(emtrends(MBase, pairwise ~ EBUS * Subregion, var = "Year", mode =  "df.error"))$emtrends)

upwelling.posthoc.base <- dplyr::full_join(upwelling.emt, upwelling.test, by = c("EBUS", "Subregion", "Year.trend", 'SE', 'df'))

upwelling.posthoc.base$Model = 'gls'

#extracts trends from the best model
upwelling.emt <- as.data.frame(emtrends(MFinal, pairwise ~ EBUS * Subregion, var = "Year", mode = "df.error")$emtrends)

upwelling.test <- as.data.frame(test(emtrends(MFinal, pairwise ~ EBUS * Subregion, var = "Year", mode =  "df.error"))$emtrends)

upwelling.posthoc.final <- dplyr::full_join(upwelling.emt, upwelling.test, by = c("EBUS", "Subregion", "Year.trend", 'SE', 'df'))

upwelling.posthoc.final$Model = 'gls + autocorrelation'

#join 
upwelling.posthoc <- rbind(rbind(upwelling.posthoc.base, upwelling.posthoc.final), upwelling.posthoc.lm)

#relabel EBUS factors
upwelling.posthoc$EBUS <- factor(upwelling.posthoc$EBUS, levels = c('California', 'Humboldt', 'Iberian/Canary', 'Benguela'))
upwelling.posthoc$Subregion <- factor(upwelling.posthoc$Subregion, levels = c('Equatorward', 'Central', 'Poleward'))

###plot comparing base model with autocorrelation model
#plot trends with CIs
upwelling.posthoc %>% 
  dplyr::filter(Model != 'lm') %>% 
  ggplot(aes(x = Subregion, y = Year.trend, group = EBUS, color = Model)) + 
  geom_errorbar(aes(ymin=lower.CL, ymax=upper.CL), width=.2,
               position=position_dodge(.9)) +
  geom_point() +
  facet_wrap(.~EBUS, ncol = 4) +
  geom_hline(yintercept = 0, linetype = 'dashed', color = 'grey60') +
  theme_bw() +
  ylab('Trend ± 95% CI') +
  ggtitle('Trends for Area Under the Curve Indicator')

#manually calculate slopes and intercepts
upwelling %>% 
    group_by(EBUS, Subregion) %>% 
    do({
      mod = lm(Percent ~ Year, data = .)
      data.frame(Intercept = coef(mod)[1],
                 Slope = coef(mod)[2])
    })

upwelling.posthoc %>% dplyr::filter(Model == 'lm') %>% view()
```

### Predicted

Look at the predicted model output from the final model (with variance partitioned and autocorrelation structure) vs just a basic linear regression.

```{r best-predict}
newdat <- upwelling %>% dplyr::select(c(EBUS, Subregion, Year))
newdat$Predicted <- predict(MFinal, newdat,re.form=NA)
newdat$Percent <- upwelling$Percent

#plot the data
ggplot(newdat, aes(x = Year)) + 
    geom_point(aes(y = Percent), color = 'black') +
    geom_line(aes(y = Predicted, color = EBUS), size = 1) + 
    geom_smooth(aes(y = Percent), method = 'lm', se = FALSE, 
                size = 0.5, linetype = 'dashed') +
    facet_wrap(EBUS~Subregion, ncol = 3, scales = 'free_y') + 
    ggtitle("Predicted (colored lines) vs Linear Regression 
            (blue, dashed lines) for EBUS and Subregion") + 
    theme_bw()
```

